Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-10-31 manuscript.tex
minor edit to abstract, new pdf

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that average neutral polymorphism for genes on the Y chromosome is \textasciitilde 2.1\% the value of their homologues on the recombining X chromosome, indicating a significant departure from neutrality and a mean chromosome-wide reduction of $\approx 93\%$. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that levels of neutral polymorphism on the Y can be explained by purifying selection against deleterious mutations, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromosome degeneration are characterized by strong selective interference among many linked sites (at least \textasciitilde 800kb), rather than simply occuring through neutral drift following gene silencing. 
\end{abstract}

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that average neutral polymorphism for genes on the Y chromosome is \textasciitilde 2.1\% the value of their homologues on the recombining X chromosome, indicating a significant departure from neutrality and a mean chromosome-wide reduction of \textasciitilde 93\%. We demonstrate that the magnitude of diversity loss is inconsistent with a purely neutral model that accounts for the occurrence of female-biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that levels of neutral polymorphism on the Y can be explained by purifying selection against deleterious mutations, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study is consistent with the prediction that the early stages of Y chromosome degeneration are characterized by strong selective interference among many linked sites (at least \textasciitilde 800kb), rather than simply occuring through neutral drift following gene silencing. 
\end{abstract}
